The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-07-01 Paper/jaccardbaetica.tex
más comentarios

\xavi{HE HECHO HASTA AQUÍ}

@@

The Roman conquest in \textit{Germania} dates from the end of the 1st century BC, during Caesar campaign and subsequently continued by Augusto. The advance of the Roman legions toward \textit{Gallia} opened a path through the Atlantic Ocean to \textit{Germania} \citep{remesal_annona_1986,
remesal_baetica_2002}. It is thought that this route would have been initiated previously through terrestrial transport, although recent studies question this position giving an important role to the Atlantic route for both the conquests of \textit{Germania} and those of \textit{Britannia} \citep{remesal_germn_2010,rubio-campillo_ecology_2018}.

The Roman studies about the presence of Dressel 20 amphorae in \textit{Germania} has not had the same repercussion as the rest of the province. This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}.

As a consequence, we do not the role that commerce agents in the participation of the distribution and exportation of olive oil in \textit{Germania} or, on the contrary, the Roman army was the responsible to supply the product in the own province \citep[156]{remesal_germn_2010}.

As the case of \textit{Britannia}, the consumption of olive oil in \textit{Germania} was not detected before the conquest. This is evidenced by the presence of Dressel 20 amphora in different sites close to \textit{limes}.

The presence of the Roman army encouraged the exchange in the province showed by the arrival of this product both civil settlement and military sites with a mayor concentration at German \textit{limes}. It seems that some Baetican centres would be assigned to the support of olive oil. However, this hypothesis can not be proven by the lack of archaeological sources \citep[125]{remesal_concierto}.

@@

The goal of this study is exploring the effect of the production patterns between different centres. We are especially interested in identifying links between production and consumption centres by using amphoric stamps. To do that, we use the CEIPAC database to collect stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.

Therefore, this study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance.

@@

We studied a dataset of 3798 stamps collected from different Dressel 20 amphora workshops in \textit{Baetica} province. For the analysis, we used a database compiled by CEIPAC group 
\citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). CEIPAC database contains over 50.000 epigraphy records found on different types of amphorae. From the database, we collected the site where stamps were found, the stamp code and the geographical coordinates.

The stamps with incomplete information were discarded and not integrated into our dataset. We finally filtered a total sample of 987 stamps composed of 130 different stamps from 81 workshops.

@@

We also created a new column with the area of the stamps. These areas, known as \textit{conventus}, were administrative centres for territorial organisation in the Roman Empire \citep[58]{ozcariz_gil_administracion_2013}.
Dressel 20 stamps were found in three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,
chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

It is important to mention that some workshops exhibited the same coordinates but they were catalogued as different workshops in Roman Studies (i.e. Tesorillo de Doña Mencia and Doña Mencia). The analysis took into account this particularity but some biases can occur due to the archaeological dataset.

We calculated the frequency of distribution in amphora stamps with the purpose of implementing an exploratory analysis (EDA) to detect the distribution and the number of stamps per each centre.

The frequency of stamps per workshop can be seen in Fig.~\ref{stamps}. Most 
workshops contained only one stamp whereas one workshop (La Catria) concentrated a high frequency of amphoric stamps with a total number of 228 stamps. The type of distribution is also frequent in amphora production where we observed a self-organised complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequencystamp.pdf}
\caption{Histogram on a log scale with base 10. X axis is represented by the number of stamps and Y axis is the frequency of workshop. The distribution is widely diverse with most workshops have only one stamp}
\label{stamps}
\end{figure}

@@

The distribution of amphora stamps in different \textit{conventus} can be seen in Fig. \ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly, workshops show a homogeneity on the frequency of stamps except for both La Catria and Arva that exhibit a big amount of stamps with 29 different code stamps. According to previous studies, those workshops became in most important centres of amphora production although they could have more intensely prospected than others \citep{arva_1997}.

@@

%\xavi{Esta estructura no me convence...tenemos que pensar algo mejor, porque ahora es: 1. hablo de britania/germania, 2. hablo de sellos en general, 3. hablo de sellos en britania/germania. Quizás juntar esto con lo de los casos de estudio, o quizás hablar de los 2 casos de estudio en la intro, y dividir la sección anterior en 2 (lo más general a la intro, lo más concreto aquí)}

We analysed a dataset of 4271 stamps  from different centres in \textit{Britannia} (2219 stamps) and \textit{Germania} (2052 stamps).
Both \textit{Germania} and \textit{Britannia} were analysed as regions and not as Roman provinces. This means that we selected centres beyond the borders of the provinces. We analyse the two provinces (Ulterior and Citerior) close to \textit{limes} in \textit{Germania} while the analysis for Britannia was extended to \textit{Caledonia}.

We used the same CEIPAC database as previous before to select amphora stamps, centres and geographical coordinates. The database contained anomalies that were eliminated using the same criteria cited above. We selected the centres with more or equal than five stamps.

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps composed of 968 different stamps from 46 centres. The centres located in \textit{Britannia} can be seen in Fig.\ref{britannia}.

@@

In \textit{Germania}, we finally collected a total of 1621 stamps consisted of 850 different stamps from 46 sites (see Fig. \ref{germania}).

@@

The approach proposed here is based on the idea of measuring the similarity between amphora workshops by quantifying similar stamps. A measure of dissimilarity has been chosen to analyse the dataset. We use the statistical technique Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was performed to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index can be expressed considering 0 as the total presence of similarity of stamps and 1 a total dissimilarity between stamps. In our case, it will be calculated the number of times that one stamp appears in an amphora workshop. This method allows to bear in mind a similar number of times for each repeated stamp per workshop. If two workshops have similar stamp codes, then the probability would be 0 whereas stamps codes are totally different when the results would be 1.

@@

Morisita-Horn index resulted in a numeric matrix of dissimilarities containing the distance value between the sites and the code stamps. The method was used to group sites with similar stamps codes. The matrix was then computed using hierarchical clustering. The algorithm was selected to cluster similar groups in order to analyse the relationship between groups of sites and the distribution of similar stamp codes. The results were visualised using a dendrogram to detect groups of sites sharing similar stamp codes.

>>>>

@@

Julius Caesar's campaigns in \textit{Gallia} during 1st BC opened the door to the invasion of \textit{Germania} during Augustus rule~\citep{remesal_annona_1986,remesal_baetica_2002}\xavi{no hay un resumen de la conquista similar para Britannia, queda raro}. Previous studies suggested that the supply to the German limes was mainly based on riverine transport, but recent works suggest that the maritime route through the Atlantic Ocean could have been more important than expected (both for Germania and Britannia).~\citep{remesal_germn_2010,rubio-campillo_ecology_2018}.\xavi{lo del océano atlántico está un poco descontextualizado; no sería mejor ponerlo al final de esta sección?}

Roman studies exploring the presence of Dressel 20 amphorae in \textit{Germania} has not had the same impact as the rest of the province\xavi{no entiendo esta frase. Que qué provincia? Germania son 2 provincias no?}. This could be explained due to the lack of archaeological sources \citep{horacio2010llegada}. As a consequence, it is still unknown if trade agents participated on the distribution of Baetican olive oil in \textit{Germania} or the supply was exclusively focused on the Roman army garrisons\citep[156]{remesal_germn_2010}\xavi{quieres decir que aquí sólo encontramos dressel 20 en yacimientos militares porque no se han excavado muchos civiles?}.

\textit{Germania} presents a similar introduction pattern of olive oil than the one discussed for \textit{Britannia}. There is no archaeological evidence for olive oil consumption before the arrival of the Roman army to the region while a majority of recovered Dressel 20 amphorae are located in the military sites that formed the German \textit{limes}\xavi{cita?}.

The presence of the Roman army encouraged the exchange in the province as shown by the arrival of this product both civil settlement and military sites with a mayor concentration at the German \textit{limes}\xavi{pero justo antes decías que no había en civiles no?}. It seems that some Baetican centres would be assigned to the support of olive oil\xavi{the support of olive oil no se entiende. Además, falta alguna cita y desarrollar esta idea}. However, this hypothesis is hard to assess given the current lack of archaeological evidence \xavi{related to BLABLABLA. Habría que detallar un poco porque hay un montón de evidencia arqueológica del limes, así que qué es exactamente lo que falta?}\citep[125]{remesal_concierto}.

@@

The goal of this study is exploring the effect of the production patterns between different centres. We are especially interested in identifying links between production and consumption centres by using amphoric stamps. To do that, we use the CEIPAC database to collect stamps from different places. The CEIPAC dataset contains over 50.000 of epigraphy records found in amphorae, mostly from \textit{Monte Testaccio}.
This study proposes a robust baseline to explore the distribution of Baetican olive oil production by computing the spatial correlation between stamps. A way to analyse is to use a quantitative framework to measure the similarity between amphoric stamps. Here we use an ecological approach based on three steps: a) to detect similarities between stamp codes, b) to explore a potential spatial correlation and c) to establish a correlation between similarity of stamps and spatial distance. 
\xavi{estos 2 párrafos explican objetivos, que deberían ir en la intro y no aquí, donde deberías empezar a hablar directamente de la BBDD}

@@

To study the stamps found in the Baetican workshops we used the data collected in the CEIPAC database of amphoric epigraphy~\citep{remesal_centro_2015} (see the database here \url{http://romanopendata.eu}). This database contains over 50.000 epigraphy records found on different types of amphorae\xavi{podrías mover esta parte a los 2 párrafos anteriores y explicar la bbdd brevemente antes de empezar a hablar de la Bética (porque es info que también aplica a los otros 2 casos de estudio no?}.

There databased allowed us to retrieve 3798 stamps found in Dressel 20 amphorae recovered from workshops within \textit{Baetica} province. Each of these stamps ha detailed records on the site where it was found, the inscribed code as well as its spatial coordinates. The stamps with incomplete information were discarded, thus finishing with a sample of 987 stamps from 81 sites and displaying 130 different code.

@@

Our dataset also contained a new categorical variable defining the \textit{conventus} of each site. The \textit{conventus} were administrative centres for territorial organisation in the Roman Empire \citep[58]{ozcariz_gil_administracion_2013}\xavi{estaría bien ampliar un poco esto; qué organizaban exactamente?}. The production area for Dressel 20 amphorae extended across three different \textit{conventus}: \textit{Hispalensis} (currently Seville, hereafter Hispalis), \textit{Cordubensis} (currently C\'ordoba, hereafter Corduba) and \textit{Astigi} (currently Écija, Sevilla, hereafter Astigi) \citep{rodriguez_economioleicola_1977,chicdatos2001,berni_millet_epigrafianforica_2008} .

@@

It is important to mention that some workshops exhibited the same coordinates but they were catalogued as different workshops in Roman Studies (i.e. Tesorillo de Doña Mencia and Doña Mencia). The analysis took into account this particularity but some biases can occur due to the archaeological dataset\xavi{No se entiende; deberías explicarlo un poco mejor y citar algo si dices eso. Piensa que la mayoría de lectores no sabrán quién es Doña Mencia}.

The first step was to compute the frequency distribution of stamp codes and analyse it using Exploratory Data Analysis (EDA). This would allow us to study the distribution of the stamps across centres as can be seen in Figure~\ref{stamps}. Most workshops contained only one stamp~\xavi{el gráfico es dificil de entender...yo leo que hay unos 20 workshops con 1 solo stamp, por lo que entiendo que no es cierto que la mayoría solo tengan 1 sello (hay como 50 workshops con entre 2 y 30 sellos no?)} One workshop (La Catria) concentrated a large percentage of the amphoric stamps with a total number of 228 stamps. The type of distribution is also frequent in amphora production where we observed a self-organised complex system pattern with a major concentration of the number of stamps in a few workshops \citep{bayesian_2018,coto-sarmiento_identifying_2018}\xavi{no lo veo; recuerda que para el rollo bayesiano tenías una línea recta al hacedr el log de los 2 ejes, no solo 1, y este plot no está construído así no?}.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/frequencystamp.pdf}
\caption{Histogram on a log scale with base 10. X axis is represented by the number of stamps and Y axis is the frequency of workshop. The distribution is widely diverse with most workshops having only one stamp}
\label{stamps}
\end{figure}

@@

The distribution of amphora stamps for each \textit{conventus} can be seen in Figure~\ref{frequency}. The majority of stamps are concentrated in \textit{Hispalis} (574 stamps) while \textit{Corduba} and \textit{Astigi} have roughly half this sample size (267 and 146 stamps). The workshops of the three conventus show a similar pattern on the stamp frequency distribution with the exceptions of 2 large workshops in \textit{Hispalis} (La Catria and Arva). On these two workshops a comparatively large amount of stamps were found (29 different code stamps on each of them)\xavi{una cosa, aquí has pasado de sellos a códigos, y los valores son distintos que en el gráfico anterior. Deberías clarificarlo antes de empezar a hablar de la figura}. According to previous studies those workshops became the most important centres of amphora production in the region\xavi{cita de estos estudios}. It is worth mentioning that a majority of these stamps were collected during ield surveys with no excavation, so this difference in sample size could be biased due to differences in intensity across the different sites~\citep{arva_1997}.

@@

%\xavi{Esta estructura no me convence...tenemos que pensar algo mejor, porque ahora es: 1. hablo de britania/germania, 2. hablo de sellos en general, 3. hablo de sellos en britania/germania. Quizás juntar esto con lo de los casos de estudio, o quizás hablar de los 2 casos de estudio en la intro, y dividir la sección anterior en 2 (lo más general a la intro, lo más concreto aquí)}
The analysis of stamps in consumption areas also used the CEIPAC database and filtered in the same way than the previous dataset. We selected the centres with more or equal than five stamps\xavi{pero esto no se hizo en el de Baetica no? Por qué?}.

The output was a dataset of 4271 stamps found in sites belonging to \textit{Britannia} (2219 stamps) and \textit{Germania} (2052 stamps). Both \textit{Germania} and \textit{Britannia} were analysed as borders and not as Roman provinces. This means that we included in the sample some centres that are actually located beyond the administrative boundaries but were considered part of the same border. Specifically, for the German \textit{liomes} we included sites from the two provinces (Ulterior and Citerior\xavi{no es superior e inferior?}) while the analysis of Britannia extended to \textit{Caledonia}.

%We analysed a dataset of 2219 stamps from different centres in \textit{Britannia}. 
%CITAR (Callender, 1965; Carreras Monfort y Funari,1998; Ayllón-Martı́ et al., 2018). 
In \textit{Britannia}, we studied a total sample of 1765 stamps from 46 sites displaying 968 unique codes\xavi{pero esto no concuerda con las 2219 de antes...es por descartar los inciertos? Hay que dejarlo claro}. The centres located in \textit{Britannia} can be seen in Figure~\ref{britannia}.

@@

In \textit{Germania}, we collected a total of 1621 stamps from 46 sites displaying 850 different stamp codes (see Figure~\ref{germania}).

@@

The approach proposed here aimed at exploring links between production and consumption centers through the identification of common amphorae codes found in different sites. For this reason we measured the similarity between amphora workshops and/or consumption sites by quantifying a pairwise distance or dissimilarity index between places (i.e. to what extent the stamp codes found on these sites were different). The chosen dissimilarity measure was the Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was applied to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}\xavi{citar reforzar más esta idea}.

@@

Considering our dataset as a non-uniform sample, this method provides a useful tool to handle large samples with different sizes and diversity \citep{wolda_similarity_1981}. Morisita-Horn index gives a value between 0 (sites have an exact stamp codes and frequency distribution) and 1 (complete difference between stamp codes). To apply Morisita-Horn we firstly calculated the number of times that each stamp code appeared in an amphora workshop. This method allowed us to bear in mind a similar number of times for each repeated stamp per workshop\xavi{no entiendo esta frase}. If two workshops hhad similar stamp code distribution then the distance index would be close to 0 whereas sites with completely unrelated stamp codes would give a distance close to 1.

@@

The Morisita-Horn index was used to generate a dissimilarity matrix containing the distance value between each pairwise site based on their code stamps. Hierarchical clustering was applied to this matrix in order to group sites with similar stamps codes distributions. The algorithm was selected to cluster similar groups in order to analyse the relationship between groups of sites and the distribution of similar stamp codes. The results were visualised using a dendrogram to detect groups of sites sharing similar stamp codes.

\xavi{hasta aquí}
